Methods for Large Scale Learning to Rank

### 2014-03-07 report/Chapters/Introduction.tex
verduidelijking Thesis Overview
structuur zonder parts

%********************************************************************
% Appendix
%*******************************************************
% If problems with the headers: get headings in appendix etc. right
%\markboth{\spacedlowsmallcaps{Appendix}}{\spacedlowsmallcaps{Appendix}}
\chapter{Motivation and Problem Statement}
Ranking is a core problem in the field of information retrieval. The ranking task in information retrieval entails the ranking of candidate documents according to their relevance for a given query. Ranking has become a vital part of web search, where commercial search engines help users find their need in the extremely large collection of the World Wide Web. One can find useful applications of ranking in many application domains outside web search as well. For example, it plays a vital role in amongst others: automatic document summarisation, machine translation, drug discovery and in determining the ideal order of maintenance operations \cite{Rudin2009}. In addition, the ranking task has been argued to be more fitting to recommender systems than rating prediction on a continuous scale \cite{Adomavicius2005,McNee2006}.\\

@@

The appearance of benchmark datasets gave insight in the performance of different Learning-to-Rank approaches, which resulted in increasing popularity of those methods that showed to perform well on one or more benchmark datasets. Up to now it remains unknown whether popular existing Learning-to-Rank methods scale well when they are used in a parallel manner using the MapReduce framework. This thesis aims to be an exploratory start in this little researched area of parallel Learning-to-Rank. A more extensive overview of my research goals and questions are described in chapter \ref{chap:goals}.\\

\chapter{Research Goals}
\label{chap:goals}
The objective of this thesis is to explore the speed-up in execution time of Learning-to-Rank algorithms through parallelisation using the MapReduce framework.
This work focuses on those Learning-to-Rank algorithms that have shown leading performance on relevant benchmark datasets.
This thesis addresses the following research questions:
\begin{description}
\item[RQ1] What are the best performing Learning-to-Rank algorithms in terms of accuracy on relevant benchmark datasets?\\

@@

\chapter{Approach}
A literature study will be performed to get insight in relevant existing techniques for large scale Learning-to-Rank. The literature study will be performed by using the following query:
\begin{itemize}
\item ("learning to rank" OR "learning-to-rank" OR "machine learned ranking") AND ("parallel" OR "distributed")
\end{itemize}
and the following bibliographic databases:
\begin{itemize}
\item Scopus
\item Web of Science
\end{itemize}

@@

\chapter{Thesis Overview}

\begin{description}
\item[Part II: Background]{introduces the reader to the basic principles and recent work in the fields of Learning-to-Rank.}
\item[Part III: Related Work]{concisely describes existing work in the field of parallel machine learning and parallel Learning-to-Rank.}
\item[Part IV: Benchmark Results]{sketches the accuracy of existing Learning-to-Rank methods on several benchmark datasets and describes the selection of Learning-to-Rank methods for the parallelisation experiments.}
\item[Part V: Selected Learning-to-Rank Methods]{describes the algorithms and details of the selected Learning-to-Rank methods.}
\item[Part VI: Implementation]{describes implementation details of the Learning-to-Rank algorithms in the Hadoop framework.}
\item[Part VII: Results \& Discussion]{presents and discusses speed-up results for the implemented Learning-to-Rank methods.}
\item[Part VIII: Conclusion]{summarizes the results and answers our research questions based on the results. The limitations of our research as well as future research directions in the field are mentioned here.}
\end{description}

>>>>

%********************************************************************
% Appendix
%*******************************************************
% If problems with the headers: get headings in appendix etc. right
%\markboth{\spacedlowsmallcaps{Appendix}}{\spacedlowsmallcaps{Appendix}}
\section{Motivation and Problem Statement}
Ranking is a core problem in the field of information retrieval. The ranking task in information retrieval entails the ranking of candidate documents according to their relevance for a given query. Ranking has become a vital part of web search, where commercial search engines help users find their need in the extremely large collection of the World Wide Web. One can find useful applications of ranking in many application domains outside web search as well. For example, it plays a vital role in amongst others: automatic document summarisation, machine translation, drug discovery and in determining the ideal order of maintenance operations \cite{Rudin2009}. In addition, the ranking task has been argued to be more fitting to recommender systems than rating prediction on a continuous scale \cite{Adomavicius2005,McNee2006}.\\

@@

The appearance of benchmark datasets gave insight in the performance of different Learning-to-Rank approaches, which resulted in increasing popularity of those methods that showed to perform well on one or more benchmark datasets. Up to now it remains unknown whether popular existing Learning-to-Rank methods scale well when they are used in a parallel manner using the MapReduce framework. This thesis aims to be an exploratory start in this little researched area of parallel Learning-to-Rank. A more extensive overview of my research goals and questions are described in section \ref{sec:goals}.\\

\section{Research Goals}
\label{sec:goals}
The objective of this thesis is to explore the speed-up in execution time of Learning-to-Rank algorithms through parallelisation using the MapReduce framework.
This work focuses on those Learning-to-Rank algorithms that have shown leading performance on relevant benchmark datasets.
This thesis addresses the following research questions:
\begin{description}
\item[RQ1] What are the best performing Learning-to-Rank algorithms in terms of accuracy on relevant benchmark datasets?\\

@@

\section{Approach}
A literature study will be performed to get insight in relevant existing techniques for large scale Learning-to-Rank. The literature study will be performed by using the following query:
\begin{itemize}
\item ("learning to rank" OR "learning-to-rank" OR "machine learned ranking") AND ("parallel" OR "distributed")
\end{itemize}
and the following bibliographic databases:
\begin{itemize}
\item Scopus
\item Web of Science
\end{itemize}

@@

\section{Thesis Overview}

\begin{description}
\item[Chapter II: Background]{introduces the reader to the basic principles and recent work in the fields of Learning-to-Rank.}
\item[Chapter III: Related Work]{concisely describes existing work in the field of parallel machine learning and parallel Learning-to-Rank.}
\item[Chapter IV: Benchmark Results]{sketches the accuracy of existing Learning-to-Rank methods on several benchmark datasets and describes the selection of Learning-to-Rank methods for the parallelisation experiments.}
\item[Chapter V: Selected Learning-to-Rank Methods]{describes the algorithms and details of the Learning-to-Rank methods selected in Chapter IV.}
\item[Chapter VI: Implementation]{describes implementation details of the Learning-to-Rank algorithms in the Hadoop framework.}
\item[Chapter VII: Results \& Discussion]{presents and discusses speed-up results for the implemented Learning-to-Rank methods.}
\item[Chapter VIII: Conclusion]{summarizes the results and answers our research questions based on the results. The limitations of our research as well as future research directions in the field are mentioned here.}
\end{description}

### 2014-03-07 report/RelatedWork.tex
structuur zonder parts

\chapter{Related Work}
The literature research is performed by using the bibliographic databases Scopus and Web of Science with the following search query: \emph{("learning to rank" OR "learning-to-rank" OR "machine learned ranking") AND ("parallel" OR "distributed")}. An abstract based manual filtering step is applied where I filter those results that actually use the \emph{parallel} or \emph{distributed} terms in context to the \emph{learning to rank}, \emph{learning-to-rank} or \emph{machine learned ranking}. Studies focusing on efficient query evaluation instead of efficient model training are likely to meet all criteria listed. As a last step I will filter out studies based on the whole document that only focus on efficient query evaluation and not on parallel or distributed learning of ranking functions.\\

@@

The following sections describe the literature study results found.

\section{CCRank}

@@

\section{Parallel ListNet using Spark}

@@

\section{nDCG-Annealing}

@@

\section{Low complexity Learning-to-Rank}

@@

\section{Distributed Stochastic Gradient Descent}

@@

\section{FPGA-based LambdaRank}

@@

\section{GPGPU for Learning-to-Rank}

@@

\section{BagBoo: A hybrid Bagging-the-Boosting Model}

@@

\section{Parallel Decision Trees}

@@

Tyree et al. \cite{Tyree2011} describe a way of parallelising \ac{GBDT} models for Learning-to-Rank where the boosting step is still executed sequentially, instead the construction of the regression trees themselves are parallelised. The parallel decision tree building is based on Ben-Haim and Yom-Tov's work on parallel construction of decision trees for classification \cite{Ben-Haim2010}. Decision trees are build layer-by-layer where the calculations needed for building each new layer in the tree are divided over the nodes using a master-worker paradigm. The data is partitioned of the workers, who compress their share into histograms and send these to the master. The master uses those histograms to approximate the split and build the next layer. The master than communicates this new layer to the workers who can use this new layer to compute new histograms. this process is repeated until the tree depth limit is reached. The tree construction algorithm parallelised with described master-worker method is \ac{CART} \cite{Breiman1984}. Speed-up experiments on the LETOR and the Yahoo! Learning to Rank challenge data sets were performed. This parallel \ac{CART}-tree building approach showed speed-up of up to 42x on shared memory machines and up to 25x on distributed memory machines.
\section{Deep Stacking Networks}
A \ac{DSN} is a processing architecture developed from the field of \emph{Deep Learning}, that uses the \ac{MSE} loss function that is easily optimisable. \ac{DSN} is based on the stacked generalization ensemble learning method \cite{Wolpert1992}, an approach where several learning models are stacked on top of each other in such a way that the output of a model is used as one of the input features of models higher up in the stack. Each layer in the stack learns has the task of learning the weights of the inputs of that layer. The close-form constraints between input and output weights allow the input weight matrices to be estimated in a parallel manner.
\section{Alternating Direction Method of Multipliers}
Duh et al. \cite{Duh2011} propose the use of \ac{ADMM} for the Learning-to-Rank task. \ac{ADMM} is a general optimization method that solves problems of the form
\begin{alignat*}{2}
\text{minimize }   &  f(x) + g(x) \\
\text{subject to } &  Ax + Bz = c
\end{alignat*}
by updating $x$ and $z$ in an alternating fashion. It holds the nice properties that it can be executed in parallel and that it allows for incremental model updates without full retraining. Duh et al. \cite{Duh2011} show how to use \ac{ADMM} to train a RankSVM model in parallel. Experiments showed the \ac{ADMM} implementation to achieve a 13.1x training time speed-up on 72 workers, compared to training on a single worker.\\

Another \ac{ADMM} Learning-to-Rank based approach was proposed by Boyd et al. \cite{Boyd2012}. Boyd et al. \cite{Boyd2012} implemented an \ac{ADMM} based Learning-to-Rank method in Pregel \cite{Malewicz2010}, a parallel computation framework for graph computations. No experimental results on parallelisation speed-up were reported on this Pregel based approach.
\section{Distributed Hyper-parameter Tuning using MapReduce}
Ganjisaffar et al. \cite{Ganjisaffar2011} observed that long training times are often a result of training many models for hyperparameter optimisation. Grid search is the \emph{de facto} standard of hyperparameter optimisation and is simply an exhaustive search through a manually specified subset of hyperparameter combinations. The authors show how to perform parallel grid search on MapReduce clusters, which is easy because grid search is an embarrassingly parallel method as hyperparameter combinations are mutually independent. They apply their grid search on MapReduce approach in a Learning-to-Rank setting to train a LambdaMART \cite{Wu2008} ranking model, a model that uses the Gradient Boosting \cite{Friedman2002} ensemble method combined with regression tree weak learners. Experiments showed that the solution scales linearly in number of hyperparameter combinations. However, the risk of overfitting grows overwhelmingly as the number of hyperparameter combinations grow, even when validation sets grow large.
\section{Distributed LambdaRank}
Svore and Burges \cite{Svore2010,Svore2012} designed two approaches to train the LambdaMART \cite{Wu2008} model in a distributed manner. Their first approach is applicable in case the whole training set fits in main memory on every node, in that case the tree split computations of the regressions trees in the LambdaMART tree ensemble are split and the data is not. The second approach distributes the training data and corresponding training computations and is therefore scalable to high amounts of training data. Both approaches achieves a six times speed-up over LambdaMART on 32 nodes compared to a single node.
\section{Learning-to-Rank with Bregman Divergences and Monotone Retargeting}
Acharyya et al. \cite{Acharyya2012} propose a Learning-to-Rank method searches for an order preserving transformation (\emph{monotone retargeting}) of the target scores that is easier for a regressor to fit. This approach is based on the observation that it is not necessary to fit scores exactly, since the evaluation is dependent on the order and not on the pointwise predictions themselves.\\

@@

Acharyya et al. \cite{Acharyya2012} defined a parallel algorithm that optimises a Bregman divergence function as surrogate of \ac{nDCG} that is claimed to be well suited for implementation of a \ac{GPGPU}. No experiments on speed-up have been performed.
\section{Parallel robust Learning-to-Rank}
Robust learning \cite{Huber1981} is defined as the task to lean a model in the presence of outliers. Yun et al. describe a \cite{Yun2014} robust Learning-to-Rank model called RoBiRank that has the additional advantage that it is executable in parallel. Their RoBiRank model uses parameter expansion to linearise a surrogate loss function, after which the elements of the linear combination can be divided over the available nodes. The Parameter expansion trick was proposed by Gopal and Yang \cite{Gopal2013}, who originally proposed it for multinomial logistic regression. Unfortunately, no speed-up experiments were mentioned for the RoBiRank method, since Yun et al. focussed their research more on robust ranking than on parallel ranking. The only reference to performance of RoBiRank in terms of speed is the statement that its training time on a computing cluster is comparable to the more efficient implementation by Lee and Lin \cite{Lee2014} of a linear kernel RankSVM \cite{Herbrich1999, Joachims2002}.

>>>>

\section{Literature study characteristics}
The literature research is performed by using the bibliographic databases Scopus and Web of Science with the following search query: \emph{("learning to rank" OR "learning-to-rank" OR "machine learned ranking") AND ("parallel" OR "distributed")}. An abstract based manual filtering step is applied where I filter those results that actually use the \emph{parallel} or \emph{distributed} terms in context to the \emph{learning to rank}, \emph{learning-to-rank} or \emph{machine learned ranking}. Studies focusing on efficient query evaluation instead of efficient model training are likely to meet all criteria listed. As a last step I will filter out studies based on the whole document that only focus on efficient query evaluation and not on parallel or distributed learning of ranking functions.\\

@@

Section \ref{sec:lit_descr} describe the approaches on parallel or distributed Learning-to-Rank found in literature.
\section{Literature description}
\label{sec:lit_descr}
\subsection{CCRank}

@@

\subsection{Parallel ListNet using Spark}

@@

\subsection{nDCG-Annealing}

@@

\subsection{Low complexity Learning-to-Rank}

@@

\subsection{Distributed Stochastic Gradient Descent}

@@

\subsection{FPGA-based LambdaRank}

@@

\subsection{GPGPU for Learning-to-Rank}

@@

\subsection{BagBoo: A hybrid Bagging-the-Boosting Model}

@@

\subsection{Parallel Decision Trees}

@@

Tyree et al. \cite{Tyree2011} describe a way of parallelising \ac{GBDT} models for Learning-to-Rank where the boosting step is still executed sequentially, instead the construction of the regression trees themselves are parallelised. The parallel decision tree building is based on Ben-Haim and Yom-Tov's work on parallel construction of decision trees for classification \cite{Ben-Haim2010}. Decision trees are build layer-by-layer where the calculations needed for building each new layer in the tree are divided over the nodes using a master-worker paradigm. The data is partitioned of the workers, who compress their share into histograms and send these to the master. The master uses those histograms to approximate the split and build the next layer. The master than communicates this new layer to the workers who can use this new layer to compute new histograms. this process is repeated until the tree depth limit is reached. The tree construction algorithm parallelised with described master-worker method is \ac{CART} \cite{Breiman1984}. Speed-up experiments on the LETOR and the Yahoo! Learning to Rank challenge data sets were performed. This parallel \ac{CART}-tree building approach showed speed-up of up to 42x on shared memory machines and up to 25x on distributed memory machines.
\subsection{Deep Stacking Networks}
A \ac{DSN} is a processing architecture developed from the field of \emph{Deep Learning}, that uses the \ac{MSE} loss function that is easily optimisable. \ac{DSN} is based on the stacked generalization ensemble learning method \cite{Wolpert1992}, an approach where several learning models are stacked on top of each other in such a way that the output of a model is used as one of the input features of models higher up in the stack. Each layer in the stack learns has the task of learning the weights of the inputs of that layer. The close-form constraints between input and output weights allow the input weight matrices to be estimated in a parallel manner.
\subsection{Alternating Direction Method of Multipliers}
Duh et al. \cite{Duh2011} propose the use of \ac{ADMM} for the Learning-to-Rank task. \ac{ADMM} is a general optimization method that solves problems of the form
\begin{alignat*}{2}
\text{minimize }   &  f(x) + g(x) \\
\text{subject to } &  Ax + Bz = c
\end{alignat*}
by updating $x$ and $z$ in an alternating fashion. It holds the nice properties that it can be executed in parallel and that it allows for incremental model updates without full retraining. Duh et al. \cite{Duh2011} show how to use \ac{ADMM} to train a RankSVM model in parallel. Experiments showed the \ac{ADMM} implementation to achieve a 13.1x training time speed-up on 72 workers, compared to training on a single worker.\\

Another \ac{ADMM} Learning-to-Rank based approach was proposed by Boyd et al. \cite{Boyd2012}. Boyd et al. \cite{Boyd2012} implemented an \ac{ADMM} based Learning-to-Rank method in Pregel \cite{Malewicz2010}, a parallel computation framework for graph computations. No experimental results on parallelisation speed-up were reported on this Pregel based approach.
\subsection{Distributed Hyper-parameter Tuning using MapReduce}
Ganjisaffar et al. \cite{Ganjisaffar2011} observed that long training times are often a result of training many models for hyperparameter optimisation. Grid search is the \emph{de facto} standard of hyperparameter optimisation and is simply an exhaustive search through a manually specified subset of hyperparameter combinations. The authors show how to perform parallel grid search on MapReduce clusters, which is easy because grid search is an embarrassingly parallel method as hyperparameter combinations are mutually independent. They apply their grid search on MapReduce approach in a Learning-to-Rank setting to train a LambdaMART \cite{Wu2008} ranking model, a model that uses the Gradient Boosting \cite{Friedman2002} ensemble method combined with regression tree weak learners. Experiments showed that the solution scales linearly in number of hyperparameter combinations. However, the risk of overfitting grows overwhelmingly as the number of hyperparameter combinations grow, even when validation sets grow large.
\subsection{Distributed LambdaRank}
Svore and Burges \cite{Svore2010,Svore2012} designed two approaches to train the LambdaMART \cite{Wu2008} model in a distributed manner. Their first approach is applicable in case the whole training set fits in main memory on every node, in that case the tree split computations of the regressions trees in the LambdaMART tree ensemble are split and the data is not. The second approach distributes the training data and corresponding training computations and is therefore scalable to high amounts of training data. Both approaches achieves a six times speed-up over LambdaMART on 32 nodes compared to a single node.
\subsection{Learning-to-Rank with Bregman Divergences and Monotone Retargeting}
Acharyya et al. \cite{Acharyya2012} propose a Learning-to-Rank method searches for an order preserving transformation (\emph{monotone retargeting}) of the target scores that is easier for a regressor to fit. This approach is based on the observation that it is not necessary to fit scores exactly, since the evaluation is dependent on the order and not on the pointwise predictions themselves.\\

@@

Acharyya et al. \cite{Acharyya2012} defined a parallel algorithm that optimises a Bregman divergence function as surrogate of \ac{nDCG} that is claimed to be well suited for implementation of a \ac{GPGPU}. No experiments on speed-up have been performed.
\subsection{Parallel robust Learning-to-Rank}
Robust learning \cite{Huber1981} is defined as the task to lean a model in the presence of outliers. Yun et al. describe a \cite{Yun2014} robust Learning-to-Rank model called RoBiRank that has the additional advantage that it is executable in parallel. Their RoBiRank model uses parameter expansion to linearise a surrogate loss function, after which the elements of the linear combination can be divided over the available nodes. The Parameter expansion trick was proposed by Gopal and Yang \cite{Gopal2013}, who originally proposed it for multinomial logistic regression. Unfortunately, no speed-up experiments were mentioned for the RoBiRank method, since Yun et al. focussed their research more on robust ranking than on parallel ranking. The only reference to performance of RoBiRank in terms of speed is the statement that its training time on a computing cluster is comparable to the more efficient implementation by Lee and Lin \cite{Lee2014} of a linear kernel RankSVM \cite{Herbrich1999, Joachims2002}.
